---
title: "assignment week 4"
author: "Nora"
date: "2026-03-18"
output: html_document
---


1) Find and download from ENCODE the peaks (i.e. bed-like format) for p300, H3K27ac and H3K27me3 in mouse embryonic stem cells (mESC) and in one differentiated tissue (heart or liver)
when there are replicates, we recommend using the bed file denoted as “conservative IDR thresholded peaks”)
2) Compare the signals across the two tissues, answering the following questions:
- What is the proportion of TSS that overlap H3K27ac, H3K27me3, and the combination of both, in each tissue?
- What is the relationship between the 3 signals (p300, H3K27ac, and H3K27me3), and is it the same in both tissues?
- What is the relationship between regulatory elements active in mESC and in the differentiated tissue?

For each question, provide figures and/or numbers (e.g. proportions of overlap, etc.), and write a paragraph describing your observations and what you think they mean.

```{r}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(ensembldb)
  library(GenomicRanges)
  library(epiwraps)
  library(rtracklayer)
  library(ggplot2)
})

```

### Question 1
downoading peaks for mouse embryonic stem cells (-> ES-Bruce4 cell line) and liver tissue (strain B6NCrl liver tissue postnatal (0 days))
```{r}
options(timeout=1800)
download.file(url= "https://www.encodeproject.org/files/ENCFF460EOA/@@download/ENCFF460EOA.bed.gz", destfile =   "p300_mESC.bed.gz")
peaks_p300_mESC <- rtracklayer::import("p300_mESC.bed.gz", format="narrowPeak")

download.file(url= "https://www.encodeproject.org/files/ENCFF105NKG/@@download/ENCFF105NKG.bed.gz", destfile =   "H2K27me3_mESC.bed.gz")
peaks_H3K27me3_mESC <- rtracklayer::import("H2K27me3_mESC.bed.gz", format="narrowPeak")

download.file(url= "https://www.encodeproject.org/files/ENCFF212HAG/@@download/ENCFF212HAG.bed.gz", destfile =   "H3K27ac_mESC.bed.gz")
peaks_H3K27ac_mESC <- rtracklayer::import("H3K27ac_mESC.bed.gz", format="narrowPeak")

download.file(url= "https://www.encodeproject.org/files/ENCFF458DFX/@@download/ENCFF458DFX.bed.gz", destfile =   "H3K27ac_mliver.bed.gz")
peaks_H3K27ac_mliver <- rtracklayer::import("H3K27ac_mliver.bed.gz", format="narrowPeak")

download.file(url= "https://www.encodeproject.org/files/ENCFF084WCQ/@@download/ENCFF084WCQ.bed.gz", destfile =   "H2K27me3_mliver.bed.gz")
peaks_H3K27me3_mliver <- rtracklayer::import("H2K27me3_mliver.bed.gz", format="narrowPeak")

download.file(url= "https://www.encodeproject.org/files/ENCFF326YGO/@@download/ENCFF326YGO.bed.gz", destfile =   "p300_mliver.bed.gz")
peaks_p300_mliver <- rtracklayer::import("p300_mliver.bed.gz", format="narrowPeak")
```

for mESC downloaded cell line ES-Bruce4
for EP300 chose "conservative IDR thresholded peaks"
for H3K27ac and H3K27me3 no such files, so I downloaded the "replicated peaks" from ENCODE4

### Question 2a
What is the proportion of TSS that overlap H3K27ac, H3K27me3, and the combination of both, in each tissue?
```{r}

ah <- AnnotationHub(localHub=TRUE)
ensdb <- ah[["AH89211"]]

#get tss sites
tx <- transcripts(ensdb)
tss <- resize(tx, width = 1, fix = "start")
seqlevelsStyle(tss) <- "Ensembl"

#make matrix to save overlaping proportion
cell_types <- c("mESC", "liver")
marks <- c("H3K27ac", "H3K27me3", "both")
overlap_matrix <- matrix(NA, nrow = length(cell_types), ncol = length(marks),
                         dimnames = list(cell_types, marks))
# fill matrix for mESC
seqlevelsStyle(peaks_H3K27ac_mESC) <- "Ensembl"
overlap_matrix["mESC", "H3K27ac"] <- sum(overlapsAny(tss, peaks_H3K27ac_mESC))/length(tss)

seqlevelsStyle(peaks_H3K27me3_mESC) <- "Ensembl"
overlap_matrix["mESC", "H3K27me3"] <- sum(overlapsAny(tss, peaks_H3K27me3_mESC))/length(tss)

overlap_matrix["mESC", "both"] <-  sum(overlapsAny(tss, peaks_H3K27ac_mESC) & overlapsAny(tss, peaks_H3K27me3_mESC)) / length(tss)

#fill matrix for liver
seqlevelsStyle(peaks_H3K27ac_mliver) <- "Ensembl"
overlap_matrix["liver", "H3K27ac"] <- sum(overlapsAny(tss, peaks_H3K27ac_mliver))/length(tss)

seqlevelsStyle(peaks_H3K27me3_mliver) <- "Ensembl"
overlap_matrix["liver", "H3K27me3"] <- sum(overlapsAny(tss, peaks_H3K27me3_mliver))/length(tss)

overlap_matrix["liver", "both"] <-  sum(overlapsAny(tss, peaks_H3K27ac_mliver) & overlapsAny(tss, peaks_H3K27me3_mliver)) / length(tss)

print(overlap_matrix)

```
Description:
In mESCs very few TSS are marked by H3K27me3, which subsequently explain that there's no TSS with both H3K27me3 and H3K27ac.
In the liver the proportion of TSS with H3K27ac is high with around 31% which indicates that the liver has many active promoters

### Question 2b
```{r}

# Proportion of H3K27ac peaks overlapping p300
prop_ac_p300_mESC <- sum(overlapsAny(peaks_H3K27ac_mESC, peaks_p300_mESC)) / length(peaks_H3K27ac_mESC)

# Proportion of H3K27me3 peaks overlapping p300
prop_me3_p300_mESC <- sum(overlapsAny(peaks_H3K27me3_mESC, peaks_p300_mESC)) / length(peaks_H3K27me3_mESC)

# Proportion of H3K27ac peaks overlapping H3K27me3
prop_ac_me3_mESC <- sum(overlapsAny(peaks_H3K27ac_mESC, peaks_H3K27me3_mESC)) / length(peaks_H3K27ac_mESC)

#same for liver 
prop_ac_p300_liver <- sum(overlapsAny(peaks_H3K27ac_mliver, peaks_p300_mliver)) / length(peaks_H3K27ac_mliver)
prop_me3_p300_liver <- sum(overlapsAny(peaks_H3K27me3_mliver, peaks_p300_mliver)) / length(peaks_H3K27me3_mliver)
prop_ac_me3_liver <- sum(overlapsAny(peaks_H3K27ac_mliver, peaks_H3K27me3_mliver)) / length(peaks_H3K27ac_mliver)

overlap_matrix <- matrix(NA, nrow = 2, ncol = 3,
                         dimnames = list(c("mESC","liver"),
                                         c("H3K27ac_p300", "H3K27me3_p300", "H3K27ac_H3K27me3")))

overlap_matrix["mESC", ] <- c(prop_ac_p300_mESC, prop_me3_p300_mESC, prop_ac_me3_mESC)
overlap_matrix["liver", ] <- c(prop_ac_p300_liver, prop_me3_p300_liver, prop_ac_me3_liver)

print(overlap_matrix)

```
### Question 2c
```{r}
# Initialize a 2x3 matrix:
# Rows = "retained in liver" vs "new in liver"
# Columns = marks
overlap_matrix <- matrix(NA, nrow = 2, ncol = 3,
                         dimnames = list(c("retained_in_liver", "new_in_liver"),
                                         c("H3K27ac", "H3K27me3", "p300")))

# --- H3K27ac ---
# retained: mESC peaks that overlap liver peaks
overlap_matrix["retained_in_liver","H3K27ac"] <- sum(overlapsAny(peaks_H3K27ac_mESC, peaks_H3K27ac_mliver)) / length(peaks_H3K27ac_mESC)

# new: liver peaks that did NOT overlap mESC peaks
overlap_matrix["new_in_liver","H3K27ac"] <- sum(!overlapsAny(peaks_H3K27ac_mliver, peaks_H3K27ac_mESC)) / length(peaks_H3K27ac_mliver)

# --- H3K27me3 ---
overlap_matrix["retained_in_liver","H3K27me3"] <- sum(overlapsAny(peaks_H3K27me3_mESC, peaks_H3K27me3_mliver)) / length(peaks_H3K27me3_mESC)
overlap_matrix["new_in_liver","H3K27me3"] <- sum(!overlapsAny(peaks_H3K27me3_mliver, peaks_H3K27me3_mESC)) / length(peaks_H3K27me3_mliver)

# --- p300 ---
overlap_matrix["retained_in_liver","p300"] <- sum(overlapsAny(peaks_p300_mESC, peaks_p300_mliver)) / length(peaks_p300_mESC)
overlap_matrix["new_in_liver","p300"] <- sum(!overlapsAny(peaks_p300_mliver, peaks_p300_mESC)) / length(peaks_p300_mliver)

# View matrix
overlap_matrix

```


