---
title: "Assignment week 7"
author: "Nora"
date: "2026-04-20"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(epiwraps)
  library(AnnotationHub)
  library(MotifDb)
  #library(memes)
  library(universalmotif)
  library(ensembldb)
  library(ggplot2)
  library(sechm)
  library(BSgenome.Mmusculus.UCSC.mm10)
  library(limma)
  library(betterChromVAR)
})


```


## Part 1
In the same dataset of ATAC on chr19 used in the first part of the practical, plot
1) the insertion (i.e. ‘cuts’) profile of nucleosome-free fragments and
2) the centers of nucleosome-containing fragments,
around the high-confidence motifs of a transcription factor of your choice (one figure containing the
two signals).


### get sites with motif for transcription factor ATF3
```{r}
# get motif of ATF3
motifs <- MotifDb::query(MotifDb, c("ATF3", "Mus"))
motif <- motifs[["Mmusculus-HOCOMOCOv10-ATF3_MOUSE.H10MO.A"]]
motif2 <- convert_motifs(motif, class="TFBSTools-PWMatrix")

#
ah <- AnnotationHub(localHub=TRUE)
genome <- ah[["AH68356"]]
# get the sequence for chr19:
chr19 <- import(genome)["19"]

# get regions with high-confidence motifs on chr19
moi <- motifmatchr::matchMotifs(motif2, chr19, out="positions", p.cutoff=1e-5)[[1]]
# convert to GRanges
moi <- as(setNames(moi,names(chr19)), "GRanges")

```

### generate tracks

```{r, eval=TRUE}
# download bam file with ATAC data
download.file("https://ethz-ins.org/content/w7/atac.chr19.bam", "atac.chr19.bam", mode="wb")
download.file("https://ethz-ins.org/content/w7/atac.chr19.bam.bai", "atac.chr19.bam.bai", mode="wb")
bam <- "atac.chr19.bam"

# tracks for cuts of nucleosome-free fragments
epiwraps::bam2bw(bam, output_bw = "nf_cuts_a.bw", paired=TRUE, binWidth=1L, type="ends", minFragLength=30,
                 maxFragLength=120, trim=4L, extend=3L, forceSeqlevelsStyle = "Ensembl")

# tracks for centers of nucleosome containing fragments
bam2bw(bam, output_bw = "nucl_centers.bw", paired=TRUE, binWidth=5L, minFragLength=140, trim=4L,
       maxFragLength=600, type="center", extend=10L, forceSeqlevelsStyle = "Ensembl")

```
### plot profile around motifs
```{r}
# list of tracks
tracks <- list.files(pattern="bw$")
names(tracks) <- gsub("\\.bw","",basename(tracks))

# extract signals around the motif occurences
# we zoom in to 300bp around the motif centers, in windows of 5bp
sm <- signal2Matrix(tracks, moi, w=5, extend=300)

# plot the signals in heatmap:
# plotEnrichedHeatmaps(sm, trim=0.95, multiScale = TRUE, axis_name = "0")

# plot aggregated signals
d <- meltSignals(sm[,c("nf_cuts_a", "nucl_centers")])
ggplot(d, aes(position, mean, colour=sample)) + geom_line(size=1.2)
```




## Part 2
In the peak counts dataset from stressed brains that we used in the second part of the practical,
identify motifs that change in accessibility between males and females, and plot a heatmap of the
deviations for the top few candidates motifs.

### download data set
```{r, eval=TRUE}
download.file("https://ethz-ins.org/content/mouse_mm38_hippocampus.peakCounts.SE.rds", "peakCounts.rds", mode="wb")

se <- readRDS("peakCounts.rds")
```

### Annotate regions

```{r}
# we restrict the regions to standard chromosomes and change the seq level style to match the genome
se <- GenomeInfoDb::keepStandardChromosomes(se, pruning.mode = "coarse")
seqlevelsStyle(se) <- "UCSC"
se <- addGCBias(se, genome=BSgenome.Mmusculus.UCSC.mm10)
```

#### get motifs
```{r}
# download motifs (archetypes)
download.file("https://hocomoco14.autosome.org/final_bundle/hocomoco14/H14CORE-CLUSTERED/H14CORE-CLUSTERED_pwm.tar.gz", "H14CORE_clusters.tar.gz")
untar("H14CORE_clusters.tar.gz")
lf <- list.files("H14CORE-CLUSTERED/pwm/", full=TRUE)
archetypes_motifs <- lapply(lf, read_matrix, headers = ">", positions = "rows")

# convert motifs for motifmatchr
archetypes_motifs <- do.call(TFBSTools::PWMatrixList, convert_motifs(archetypes_motifs, class="TFBSTools-PWMatrix"))
names(archetypes_motifs) <- make.unique(sapply(archetypes_motifs, \(x) gsub("\\..+","",x@name)))
```

#### scan motif matches
```{r}
matches <- motifmatchr::matchMotifs(archetypes_motifs, se, genome=BSgenome.Mmusculus.UCSC.mm10)
```



### Differential analysis with deviation
```{r}
# get motif deviations
dev <- betterChromVAR(se, matches)

# name groups
dev$stress <- ifelse(grepl("FSS", colnames(dev)), "Stressed", "Control")
dev$sex <- ifelse(grepl("female", colnames(dev), ignore.case = TRUE), 
                  "Female", "Male")

design <- model.matrix(~stress + sex, data=as.data.frame(colData(dev)))

fit <- eBayes(lmFit(assay(dev, "z"), design))
res <- as.data.frame(topTable(fit, coef = "sexMale" , number = Inf))
res$motif <- row.names(res)
# ggplot(res, aes(logFC, -log10(adj.P.Val), label=motif)) + geom_text()

# visualize top hits as heatmap
sechm(dev, head(row.names(res)), gaps_at="sex", assayName = "z")
```


