---
title: "Assignment week 9"
author: "Nora"
date: "2026-05-12"
output: html_document
---

```{r}
suppressPackageStartupMessages({
  library(GenomicRanges)
  library(epiwraps)
  library(ggplot2)
  library(rGREAT) # Gene Ontology enrichment among genomic regions
})
```



## Part1 
Download and decompress the following archive:
https://ethz-ins.org/content/clustering_assignment.zip
This contains the bigwig files and peaks (bed) files for three TFs of the CREB family (all restricted to chr1; aligned against the hg38 genome)

#### downloading data
```{r, eval=FALSE}
options(timeout = 6000)
download.file("https://ethz-ins.org/content/clustering_assignment.zip", "clustering_assignment.zip")
unzip("clustering_assignment.zip")
list.files()
```

#### Prepare the regions and the tracks

```{r}
tracks <- list.files(pattern="bw$")

### To get a clearer picture, focus on high-confidence peaks from each factor to define the universe of regions
peaks <- list.files(pattern="bed$")
# we first import the peaks
peaks <- lapply(peaks, rtracklayer::import.bed)
# we'll focus on the high-quality peaks
peaks <- lapply(peaks, FUN=function(x) x[x$score>800])
# we get the union of non-redundant regions
regions <- reduce(unlist(GRangesList(peaks)))

```


## Part 2 Clustering and visualization
Use clustering and visualization to illustrate the relationship between the binding of the different proteins
```{r}

ese <- signal2Matrix(tracks, regions, extend=2000)
set.seed(123)  # to ensure that it gives the same results everytime

# cluster version 1
cl <- clusterSignalMatrices(ese, k=4)
rowData(ese)$cluster <- cl # put cluster as rowData for later maipulation of the object
head(rowData(ese))
# plotting heatmap with defined colours
mycolors <- c("1"="red", "2"="blue", "3"="darkgreen", "4"="black", "5"="purple")
plotEnrichedHeatmaps(ese, row_split=rowData(ese)$cluster, mean_color=mycolors, colors=c("white","darkred"), use_raster = FALSE)


# Clustering using relative signal:
set.seed(123)
cl <- clusterSignalMatrices(ese, k=5, scaleRows = TRUE)
rowData(ese)$cluster <- cl # put cluster as rowData for later maipulation of the object
d <- meltSignals(ese, splitBy=cl)
ggplot(d, aes(position, mean, colour=sample)) + geom_line() + facet_wrap(~split)
plotEnrichedHeatmaps(ese, row_split = cl, scale_rows = "global", mean_color=mycolors, use_raster = FALSE)
```



## Part 3 Enrichement analysis
Use enrichment analysis (either GO or motif) on at least one of the clusters
Write a paragraph describing your results
Save your assignment in a R markdown named assignment.Rmd, render it, and push the html file to this folder in your github repository


Find what's enriched in cluster 2 with respect to the clusters:
```{r}
# splitting the regions by cluster:
split_regions <- split(rowRanges(ese), rowData(ese)$cluster)
lengths(split_regions)

res <- great(split_regions[["2"]], gene_sets="GO:BP", tss_source="hg38", 
             background=regions, cores=2) # background is set to regions -> cluster 2 is compared with the other clusters

bp <- getEnrichmentTables(res)

head(bp)

```

We plot the top Biological Processes:
```{r, fig.width=9, fig.height=6}
ggplot(head(bp,15), aes(fold_enrichment, reorder(description, p_adjust), 
                        size=observed_region_hits, color=-log10(p_adjust))) + 
  geom_point() + scale_color_viridis_c()
```
The background for calculating the enrichment was set to the regions. So we can check the enrichment of GO terms of cluster 2 compared to the remaining 4 clusters. 
From the plot before we know that in cluster 2 there's a high level of Creb1 binding but low binding of the other two proteins (Creb3 & Creb3L1).
Multiple GO terms are related to the nervous system, which is not surprising as Creb1 is known to be important for learning/memory and neural function. In addition there are 3 GO terms with a high fold enrichment but low observed region hits and therefore not surprising also a low p-value: myofibril assembly, endocrine system development, protein localization to cell surface.
The GO term "central neervous system neuron differentiation" shows a high fold-enrichment and at the same time also a low p-value (eventhough the observed region hits is not high.) So in a next step I would be interesting to take a closer look at that GO-term.






