---
title: "Assignement week 6"
author: "Nora"
date: "2026-03-30"
output: html_document
---


```{r setup}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(GenomicRanges)
  library(ggplot2)
  library(motifmatchr) # for scanning sequences for matches of given motifs
  library(Biostrings) # for handling sequences
  library(MotifDb) # database of motifs
  library(TFBSTools) # for handling some motif formats
  library(universalmotif) # for converting motifs to various formats
  library(PWMEnrich) # for R-based motif enrichment analysis
  library(GenomeInfoDb)
  library(S4Vectors)
})
```

Choose a transcription factor, e.g. CREB1, REST, GATA5, EGR1, GCR (or any of your choice that has a motif and available ChIPseq data)
Download the (e.g. Mouse) peaks for that factor (in whatever cell type)
Identify the instances of the factor's motif
Answer the following questions:
Of all the peaks, what proportion contains a motif for the factor?
Expected form of an answer: of the XX peaks, XX (XX%) contain a motif
Of all instances of that motif in the genome (or in one chromosome), what proportion is bound by the factor (i.e. has a peak)?
Expected form of an answer: of the XX motif instances, XX (XX%) overlap a peak
Don’t forget to render your markdown and push it as assignment.html !


## downloading peaks for transcription factor REST in myocytes (mouse)

```{r}
download.file("https://www.encodeproject.org/files/ENCFF368VWJ/@@download/ENCFF368VWJ.bed.gz", "ENCFF368VWJ.bed.gz")
peaks <- rtracklayer::import("ENCFF368VWJ.bed.gz", format="NarrowPeak")
seqlevelsStyle(peaks) <- "Ensembl"  # change the convention of the chromosome names to ensembl 

ah <- AnnotationHub()
genome <- ah[["AH68356"]]
genome_seqs <- import(genome)
# save genome as fasta file
Biostrings::writeXStringSet(genome_seqs, "genome.fa")

```

## get motif of REST

```{r}
# we search for "CTCF" in the motif database
motifs <- query(MotifDb, c("REST"))
# there are several matching motifs:
names(motifs)
# choose mouse motif
motif <- motifs[["Mmusculus-HOCOMOCOv10-REST_MOUSE.H10MO.A"]]
```
## Motif scanning
### Question 1: Of all the peaks, what proportion contains a motif for the factor?
```{r}
# convert format of motif 
motif2 <- convert_motifs(motif, class="TFBSTools-PWMatrix")

# function returns position of each motive match with peaks
moi_p <- motifmatchr::matchMotifs(motif2, subject=peaks, genome=Rsamtools::FaFile("genome.fa"),
                                out="positions")
#hist(moi@unlistData@elementMetadata@listData$score)

# which peaks contain a motiv
table_motiv_for_factor <- table(overlapsAny(query = peaks, subject = moi_p))
n_peaks <- length(peaks)
#n_peaks <- sum(table_motiv_for_factor) returns the same number
cat("of the", n_peaks, "peaks,", table_motiv_for_factor[2], "(", table_motiv_for_factor[2]/n_peaks*100, "%) contain a motiv")

```



### Question 2: Of all instances of that motif in the genome (or in one chromosome), what proportion is bound by the factor (i.e. has a peak)?

```{r}
moi_g <- motifmatchr::matchMotifs(motif2,
                                subject=genome_seqs,
                                genome=Rsamtools::FaFile("genome.fa"),
                                out="positions")

total_genome_motifs <- length(moi_g[[1]]@unlistData)
motifs_in_peaks <- length(moi_p@unlistData)

cat("Of the", total_genome_motifs, "motif occurrences,", motifs_in_peaks, 
    "(", motifs_in_peaks/total_genome_motifs*100, "%) are bound by peaks\n")

```



